Probabilistic Catalogs for Crowded Stellar Fields

### 2012-11-19 paper/ms.tex
Working on the introduction

\section{Introduction}
Traditional practice in astronomy is to take images of the sky, detect
or enumerate sources visible in those images, and create catalogs.
These catalogs are then used to perform the fundamental astronomical
measurements, which might include something like the three-dimensional
structure of the Galaxy or the two-point correlation function of
galaxies.  Indeed, the process of catalog construction is so ``baked
in'' to our ideas about what astronomy is, we sometimes forget that
the catalog is \emph{not} the fundamental data product of astronomy; catalogs are produced
from imaging; their production involves many decisions and ideas
that go way beyond the information provided to the telescope by the
incident intensity field. In addition, catalogs are not usually the final
\emph{goal} of any imaging project or survey. Typically, they are produced in
order to facilitate the scientific study of populations of objects (e.g. the
initial mass function of a population of stars), or to provide a sky-search
capability to the community who might be interested in only a small subset of
objects. Standard tools for generating catalogs from astronomical imaging
include \sex~\citep{sextractor}, DAOPHOT \citep{1987PASP...99..191S}, and SDSS
PHOTO \citep{photo}.

Telescopes don't make catalogs \citep{2011EAS....45..351H}, they
measure the intensity field.  Viewed through the lens of probabilistic
inference, the goals of astronomy are to take the information in the
telescope-generated records of the intensity field and deliver it to
the quantities of astronomical interest with as little loss as
possible.  Insertion of a catalog-generation step in the inference
pipeline between the raw imaging and the final astrophysical analyses
is typically very lossy; the hard decisions of catalog making destroy
information.  Probability theory suggests that it will be much less
``lossy'' to pass forward not a catalog but a probabilisitic
description of all the catalogs that could be consistent with the
imaging---a posterior probability in (enormously large) catalog space.
This article represents an attempt at this ambitous goal in the specific
situation where the only objects in the field are stars.

@@

Essentially, the creation of a catalog is an attempt to answer the question,
``Given the image we have obtained, what objects are present in the field and
what are their properties?'' This motivates a probabilistic (Bayesian)
approach to making catalogs. The output of such an approach would not be a
single answer to this question (i.e. a single ``point estimate'' catalog), but rather a posterior probability distribution over the space of possible catalogs.
Constructing such a probability distribution is
challenging for a number of reasons. Firstly, the number $N$ of objects in the
image (and that should be listed in the catalog) is itself unknown. Secondly, if $N$ is large, then the parameter space
of positions and properties (flux, size, etc) of the objects is also large.
This can cause Markov Chain Monte Carlo (MCMC) algorithms difficulties -- they
may take a long time to converge to the target posterior distribution over
the space of catalogs.

Bayesian object detection (as this problem is sometimes called) has been
implemented both inside and outside of astronomy
\citep[e.g.][]{object, 2011MNRAS.415.3462F}. However, the
\citet{2011MNRAS.415.3462F} approach makes the assumption of a
known number of objects $N$.
This assumption is required for the
{\tt MultiNest} sampler \citep{multinest} to be applicable.
Using the results from the known $N$
run, it is possible (under certain circumstances) to reconstruct what the
results would have been if a variable-$N$ simulation had been used. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap). What is really required is a variable
dimension model, where $N$ is an unknown quantity to be inferred from the data.
The computational implementation of these models will require tools such as
reversible jump Markov Chain Monte Carlo \citep{rjmcmc}. Other statistical
methods such as maximum likelihood have also been used to model crowded fields
\citep[e.g.][]{irwin}. However, maximum likelihood is not completely appropriate
because it may lead to overfitting. In this situation overfitting would result
in more stars being added to the model to explain small positive fluctuations
in the image which are actually due to noise.

@@

@@

>>>>

\section{Introduction}
Traditional practice in astronomy is to take images of the sky, detect
or enumerate sources visible in those images, and create catalogs.
These catalogs are then used to perform the fundamental astronomical
measurements, which might include something like the three-dimensional
structure of the Galaxy or the two-point correlation function of
galaxies.  Indeed, the process of catalog construction is so ``baked
in'' to our ideas about what astronomy is, we sometimes forget that
the catalog is \emph{not} the fundamental data product of astronomy; catalogs are produced
from imaging; their production involves many decisions and ideas
that go way beyond the information provided to the telescope by the
incident intensity field. In addition, catalogs are not usually the final
goal of any imaging project or survey. Typically, they are produced in
order to facilitate the scientific study of populations of objects (e.g. the
initial mass function of a population of stars), or to provide a sky-search
capability to the community who might be interested in only a small subset of
objects. Standard tools for generating catalogs from astronomical imaging
include \sex~\citep{sextractor}, DAOPHOT \citep{1987PASP...99..191S}, and SDSS
PHOTO \citep{photo}.

Telescopes don't make catalogs \citep{2011EAS....45..351H}, they
measure the intensity field.  Viewed through the lens of probabilistic
inference, the goals of astronomy are to take the information in the
telescope-generated records of the intensity field and deliver it to
the quantities of astronomical interest with as little loss as
possible.  Insertion of a catalog-generation step in the inference
pipeline between the raw imaging and the final astrophysical analyses
is typically very lossy; the hard decisions of catalog making destroy
information.  Probability theory suggests that it will be much less
lossy to pass forward not a catalog but a probabilisitic
description of all the catalogs that could be consistent with the
imaging---a posterior probability in (enormously large) catalog space.
This article represents an attempt at this ambitous goal in the specific
situation where the only objects in the field are stars.

@@

Essentially, the creation of a catalog is an attempt to answer the question,
``Given the image we have obtained, what objects are present in the field and
what are their properties?'' This motivates a probabilistic (Bayesian)
approach to making catalogs. The output of such an approach would not be a
single answer to this question (i.e. a single ``point estimate'' catalog), but
rather a posterior probability distribution over the space of possible catalogs.
Numerically, Bayesian Inference is often implemented using Markov Chain Monte
Carlo (MCMC) algorithms in order to generate random samples from the posterior
distribution.

Sampling a posterior probability distribution for catalogs is likely to be
challenging for a number of reasons. Firstly, the number $N$ of objects in the
image (and that should therefore be listed in the catalog) is itself unknown.
Secondly, if $N$ is large, then the parameter space
of positions and properties (flux, size, etc) of the objects is also large.
This can cause Markov Chain Monte Carlo (MCMC) algorithms difficulties -- they
may take a long time to converge to the target posterior distribution over
the space of catalogs. Thirdly, this problem is subject to the so-called
label-switching problem that is commonly encountered in mixture modeling
\citep{label_switching}. Given any proposed catalog, another catalog that is
equally plausible is the catalog obtained by shuffling the entries of the first
catalog. This leads to a posterior distribution with $N!$ identical peaks in
parameter space. This can lead to difficulties with certain (generally very
effective) MCMC algorithms such as the affine-invariant stretch move
\citep{goodman, emcee}.

Bayesian object detection (as this problem is sometimes called) has been
implemented both inside and outside of astronomy
\citep[e.g.][]{object, 2011MNRAS.415.3462F}. However, the
\citet{2011MNRAS.415.3462F} approach makes the assumption of a
known number of objects $N$.
This assumption is required for the
{\tt MultiNest} sampler \citep{multinest} to be applicable.
Using the results from the known $N$
run, it is possible (under certain circumstances) to reconstruct what the
results would have been if a variable-$N$ simulation had been used. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap). What is really required is a variable
dimension model, where $N \in \{0, 1, 2, ... \}$ is an unknown quantity to be
inferred from the data.
The computational implementation of these models will require tools such as
reversible jump Markov Chain Monte Carlo \citep{rjmcmc}. Other statistical
methods such as maximum likelihood have also been used to model crowded fields
\citep[e.g.][]{irwin}. However, maximum likelihood is not completely appropriate
because it may lead to overfitting. In this situation overfitting would result
in more stars being added to the model to explain small positive fluctuations
in the image which are actually due to noise.

@@

\bibitem[Foreman-Mackey et al.(2012)]{emcee} Foreman-Mackey, 
D., Hogg, D.~W., Lang, D., \& Goodman, J.\ 2012, emcee: The MCMC Hammer, arXiv:1202.3665

\bibitem[\protect\citeauthoryear{Goodman \& Weare}{2010}]{goodman}
Goodman, J., Weare, J., 2010, Ensemble Samplers with Affine Invariance, Comm. App. Math. Comp. Sci., 5, 6.

@@

\bibitem[Jasra et al(2005)]{label_switching} Jasra, A., Holmes, C.~C, and
Stephens, D.~A., Markov Chain Monte Carlo Methods and the Label Switching
Problem in Bayesian Mixture Modeling, Statistical Science
Vol. 20, No. 1, pp. 50-67
